---
title: "911 calls Salford"
author: "Daniel Garcia"
date: "2022-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
# importing necessary libraries 
library(tidyverse)
library(moderndive)
library(rcompanion)
library(skimr)
```
#### this project models 911 calls in the county of Montgomery, Pennsylvania. 
#### {r} [Kaggle link to data] (https://www.kaggle.com/datasets/mchirico/montcoalert)

## 911 calls
```{r}
# reading data 
df1= read_csv("C:\\Users\\dgray\\OneDrive\\data\\911.csv")
glimpse(df1)
```


```{r}
# checking range time data was taken 
range(df1$timeStamp)
```
goes from 12/10/2015 to 07/29/2020


# Homogenous method

```{r}
# count of 911 calls by twp in descending order. Lower Merion has the most
twp_df= df1 %>% 
  group_by(twp) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))
twp_df
```
### Lower Merion has the highest crime rate, so lets analyze them

```{r}
# creating dataframe with only the town Lower Merion
EG_df = df1 %>% 
  filter(twp=="EAST GREENVILLE")
EG_df
```

```{r}
# plotting cases by year in lower merion
ggplot(EG_df, mapping=aes(x=timeStamp)) +
  geom_histogram(bins=20, color="white", fill="#CC9900")
```

```{r}
glimpse(EG_df)
```

## finding lambda 
```{r}
# changing format of timestamp to discard time and only include date 
eastG_df= EG_df %>% 
  mutate(format(as.POSIXct(EG_df$timeStamp, format='%m/%d/%Y %H:%M:%S'),format='%m/%d/%Y')) %>% 
  rename("TimeStamp"= "format(...)") %>% 
  select(-timeStamp)
glimpse(eastG_df)
```

```{r}
# grouping by date and counting number of 911 calls for each day 
num_calls_per_day = eastG_df %>% 
  group_by(TimeStamp) %>% 
  summarise(count=n())
num_calls_per_day
```
 
```{r}
# calculating number of distinct days
n_days=n_unique(num_calls_per_day$TimeStamp)
n_days

# summing the number of total 911 calls
total_calls= sum(num_calls_per_day$count)
total_calls
```
 
```{r}
# Lambda is total calls divided by the number of days
lambda= total_calls/n_days
lambda
```




### visualize


```{r}
# distribution of number of 911 calls per day
ggplot(data= num_calls_per_day, mapping=aes(x=count)) +
  geom_histogram(bins=12, color="black", fill="orange") +
  labs(title="Distribution of the number of calls per day", x="Number of 911 calls")
```




## Waiting times by day
```{r}
# average waiting time is n/lambda
avg_wait_time = 1/lambda
avg_wait_time
```
```{r}
avg_wait_time_min = avg_wait_time*24*60
avg_wait_time_min
```
There will be a 911 call on average every 14 hrs and 15 minutes, 
approximately. 


```{r}
# variance of waiting time
var_wait_time = 1/(lambda**2)
var_wait_time
```
about 21 minutes^2



## simulating trajectory
```{r}
lambda= 1.685019
njumps<- 20

N<- 0:njumps

time<- c()

time[1]<- 0

set.seed(333422)

for (i in 2:(njumps+1))
time[i]<- time[i-1]+round((-1/lambda)*log(runif(1)),2)

plot(time, N, type="n", xlab="Days", ylab="Calls",
panel.first = grid())

segments(time[-length(time)], N[-length(time)],
time[-1]-0.07, N[-length(time)], lwd=2, col="Dark green")

points(time, N, pch=20, col="black")
```

## Predictions
```{r}
# lets predict the wait time (in minutes) till the 20th call
wait_20 = 20*avg_wait_time_min
wait_20
wait_20/(60*24)
```
17091.79 minutes, which is approximately 11 days, 20 hours, and 52 minutes 







## nonhomogenous poisson
```{r}
# changing time format to only include hour of the day 
df5= EG_df %>% 
  mutate(format(as.POSIXct(EG_df$timeStamp, format='%m/%d/%Y %H:%M:%S'),format='%H')) %>% 
  rename("Time"= "format(...)") %>% 
  select(-timeStamp)
glimpse(df5)
```

```{r}
# number of calls by hour of the day
num_calls_per_hr = df5 %>% 
  group_by(Time) %>% 
  summarise(count=n())

# change Time variable to numeric
num_calls_per_hr$Time=as.numeric(num_calls_per_hr$Time)
num_calls_per_hr$count=as.numeric(num_calls_per_hr$count)
glimpse(num_calls_per_hr)


```




```{r}
glimpse(num_calls_per_hr)
```

### splitting time into 4 intervals

```{r}
# LM1 is from 4AM to 9AM 
EG1 = num_calls_per_hr %>% 
  filter(Time>=0 & Time<=5)
EG1
```

```{r}
# LM2 from 10AM to 3PM
EG2 = num_calls_per_hr %>% 
  filter(Time>=6 & Time<=11)
EG2
```

```{r}
# LM3 from 4PM to 9PM
EG3 = num_calls_per_hr %>% 
  filter(Time>=12 & Time<=17)
EG3
```

```{r}
# LM4 from 10PM to 3AM 
EG4 = num_calls_per_hr %>% 
  filter(Time>=18 & Time<=23)
EG4
```

### Calculate average per interval


```{r}
# calculating intensity rate for each interval
EG1_avg = mean(EG1$count)
EG2_avg = mean(EG2$count)
EG3_avg = mean(EG3$count)
EG4_avg = mean(EG4$count)

```

```{r}
# show intensity rate for each interval 
EG1_avg
EG2_avg
EG3_avg
EG4_avg

```

### graphing intensity rate function and integrated intensity rate function 


```{r}
# plotting intensity rate function
df_ = data.frame(vec1 = c(EG1_avg, EG2_avg, EG3_avg, EG4_avg, EG4_avg),
                 len= c(0, 5, 11, 17, 23))
ggplot(df_, mapping=aes(x=len, y=vec1)) +
  geom_step(color="Purple", size=1) +
  geom_point(size=3, shape=1) +
  labs(title="Intensity rate function", y="intensity rate", x="hour of day")
```

```{r}
# plotting integrated rate function
lm = data.frame(lambda= c(0, 127.5, 476.5, 939.5, 1290.5),
                int= c(0, 5, 11, 17, 23))
ggplot(lm, aes(x=int, y=lambda)) +
  geom_line(color="orange", size=1.25) +
  geom_point() +
  labs(title="Integrated intensity rate function", x="hour of day", y="integrated rate function")
```






